Can government public communications elicit undue trust? Exploring the interaction between symbols and substantive information in communications

### 2019-04-13 paper2_JPART_rev_01.Rmd
Add files via upload

---
title: "Can government public communications elicit undue trust? Exploring the interaction between symbols and substantive information in communications"
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
csl: apa.csl
urlcolor: blue
---

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. Those studies that have explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still we lack theoretical understanding of the mechanisms underlying citizens' responses to symbolic elements, and the interaction between symbols and substantive information. Building on the social psychology Elaboration Likelihood Model I hypothesize that symbols increase citizens' trust in two ways: First, directly, by evoking positive feelings that are transferred to the organizations and policies presented in the communication. Second, indirectly, by conducing citizens to pay less attention to the substantive information in the communication. As a result, citizens disregard logically unpersuasive information that could decrease their trust. I examine these expectations via a survey experiment, focusing on the Israeli Environment Protection Ministry. The results support the hypotheses and suggest that the effect of symbolic elements is stronger when communications include logically unpersuasive information. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

One of the main remedies proposed for this is greater investment in direct government communications to citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014; @canel_2013]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant research on the effect of government public communications on citizens trust has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. This focus on information and transparency is in line with the positive normative view of government public communication and public relations, as articulated by the public administration literature on this topic, led by the works of Merdechai Lee. Communications with citizens are perceived as a means for accomplishing democratic accountability, facilitating the effective implementation of their missions, and aligning citizens' perceptions of public organizations with their actual performance [@canel_2013; @lee_2000; @lee_2012; @lee_2001; @lee_2008; @lee_2011]. Importantly, according to this view, public communications are expected to increase citizens' trust in bureaucracies mainly insofar as these communications provide citizens with favorable information on the performance of organizations and logically persuasive explanations for their policies.

Recently, however, a few studies pointed to the additional effect of *symbolic* elements that are entangled in government communications, such as  agencies' names and brand logos, figures, images and celebrity endorsements [@alonbarkat_2017; @karens_2016; @teodoro_2018]. These studies experimentally demonstrated that government organizations may cause citizens to view them more favorably and trust them through familiar and well-designed symbolic elements that evoke positive associations and emotions. Still, we lack a theory and robust empirical evidence of the exact mechanisms through which these symbols increase citizens’ trust. These studies also did not examine the interaction between the symbolic elements and the substantive content of information in public communications.

This research gap has normative significance. Whereas the responsiveness of citizens to substantive information is perceived favorably from a democratic standpoint (assuming that it is accurate), establishing trust through symbolic elements can be highly problematic. Shaping citizens' perceptions of governments by evoking associations and emotions stands against the normative expectation that citizens' judgments should be based on their critical thinking. Furthermore, we need to be aware that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017]. Therefore, this study focuses on the following questions: *What are the psychological mechanisms by which symbolic elements in government communications affect citizens' trust in government organizations and their policies? Does the effect of symbolic elements on trust depend on the substantive content of communications?* *Do symbols offset the rational, negative effect of unfavorable or logically unpersuasive information?*

To address these research questions, I draw upon the social psychology Elaboration Likelihood Model (ELM). ELM provides a comprehensive theoretical framework for various psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust in two ways: *First*, they may evoke positive feelings and associations that are transferred from the symbols to the relevant government entities and policies, and thereby increase citizens' trust in them. *Second*, symbolic elements may increase citizens' trust in government entities and policies by causing them to disregard the content of logically unpersuasive information that could otherwise decreased their trust.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry. Specifically, I examine how their trust in these public policies is affected by the symbolic elements in the communications (real symbols vs. fake symbols and no symbols conditions) as well as by the substantive information in them (logically persuasive vs. unpersuasive policy plans).

I found that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans. The latter were less inclined to think about the weak, unpersuasive content, because they were distracted by the symbols. I discuss the theoretical and normative implications of these findings.

@@

A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. Symbols are fundamental for almost any form of human interaction, and specifically in the context of politics and public affairs [@edelman_1964; @edelman_1988; @elder_1983]. Conselletions of symbols in discourse form consolidated narratives, which play an key role in shaping people's perceptions of policy problems and solutions [@miller_2012; @miller_2015].

My focus in this study is on those symbols that are strategically used by government organizations in their public communications, are designed to distinguish them and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can also be found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared citizens who were personally exposed to prolonged power outages and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations and policies associated with them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the government quality assessed.

@@

Even though these studies substantiate a link between the appearance of symbolic elements in communications and citizens' greater trust in public organizations and policies, they did not directly investigate the causal mechanisms that underlie this effect. Therefore, we still lack understanding of the circumstances under which symbols are likely to be more/less effective in shaping citizens' trust. Specifically, these studies neither theorized nor empirically tested the interaction between symbolic elements and citizens’ processing of substantive information in communications. We do not know whether well-designed symbols reinforce, or offset, citizens' rational responses to information. Specifically, as explained above, it is essential to assess the possibility that symbols can unduly compensate for unfavorable or logically unpersuasive information that should yield negative responses.

To address this theoretical gap in public administration research, I propose a nuanced theoretical framework for the effects of symbols in communications and their interaction with information, building on social psychology theory on persuasive communication.

@@

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" communications [@petty_1986; @petty_2011]. It theorizes when and how individuals are more/less likely to change their attitudes following a communication or a message. Accordingly, ELM can be applied to various settings, types of attitudes and forms of communications. It has been specifically applied to studies on persuasion or attitudinal change in various areas, including commercial advertising [e.g. @petty_1983; @dotson_2000; @wheeler_2005; @moore_1987], public health campaigns [@wilson_2007; @angst_2009; @ruiter_2001; @briol_2006], media coverage [@petty_2002; @vanderwurff_2016] and media entertainment [@young_2008; @slater_2002]. So far however the relevance of ELM has not been directly examined in relation to changes in citizens' trust in public organizations in response to their public communications.

According to ELM, the responses of individuals to persuasive communications depends on how much effort they devote to thinking about the content of the message (or "the amount of thinking" according to ELM terminology). Persuasion, or attitudinal change, can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little cognitive effort, such as an unconscious transfer of affect[^footnote_dual_process_theories]. In turn, these two mechanisms are likely to cause divergent types of persuasion. Persuasion via the central route is expected to be more enduring, resistant and predictive of behavior, compared with the peripheral route. Of course, often people are located in between the two ends of the continuum, and thus persuasion is determined by a combination of these processes.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and able to do so. More positive information about organizations' performance and more logically persuasive arguments supporting their actions would increase citizens' trust in these organizations and their actions and vice versa. Conversely, symbolic elements are potential peripheral cues. they can increase citizens' trust via the peripheral route, by evoking positive emotions and associations that are attributed to government organizations and their policies [@alonbarkat_2017]. However, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing substantive information that would otherwise decrease their trust, namely negative information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive arguments supporting their actions and policies). The latter assertion entails an interaction between symbols and the substantive information in communications: symbolic elements in public communications are more likely to affect citizens' trust when communications include negative information or weak logical arguments.[^footnote_interaction_strong_information]

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates citizens' responses to government organizations' communications about their policy plans. Such communications can be found in publicized reports, pamphlets, websites, new media pages and ads. I focus on communication of future policy plans, as opposed to communications on performance of existing policies. This allows me to utilize a real-world empirical case, yet still to manipulate the content of the communication in a reliable manner. Most citizens do not have prior knowledge of organizations' future policy plans, and will therefore be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM) that is responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in the EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans, in which I manipulated the informative and symbolic elements, are presented to the participants as sections from the EPM's official publication of its annual work plan.

@@

In the course of the survey experiment, I was able to confirm that the EPA logo and the images of the celebrities from the campaigns were familiar to the participants. A total of `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize the EPA logo to some extent, and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`%, respectively). In robust analyses I further exclude those who do not recognize these symbols.[^footnote_prior_studies]

@@

[^footnote_prior_studies]: In two prior surveys, I collected further information about the affect and symbolic associations of these symbols. These results are reported in the supplementary appendix and substantiate my empirical assumption that these elements evoke positive emotions, and that citizens automatically associate them with the EPM.

@@

In the control no-symbols condition, the two policies were displayed in a minimal black and white design, and without the logo and the celebrities’ images. In the control fake-symbols condition, the green color was replaced with blue, the EPM logo was replaced by a fake logo and the images of celebrities from the campaigns were replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition was to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which could be attached to the government ministry and environmental issues.  In that sense, the "fake" symbols could be regarded as a placebo.

@@

[^footnote_symbols_manipulation_checks]: In two prior surveys, I confirmed, empirically, that the fake symbols are unfamiliar, do not have strong positive or negative affect, and are not strongly associated with environment policy and EPM. The results of these examinations are reported in the supplementary appendix.

@@

The main independent variables are the above mentioned manipulations for information and symbols, and the interaction between them. The outcome variable is citizens' trust in policy (or the perceived trustworthiness of the policy). This variable is measured in the survey for each policy plan using a composite index of six likert scale items (Cronbach’s alpha = `r alpha_waste_trust`), which were adopted and slightly modified from validated trust scales used by previous public administration studies on trust in government [@grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @karens_2016; @porumbescu_2016; @grimmelikhuijsen_2017]. Participants were asked to indicate their agreement with the following statements, between 1 (weakly agree) and 7(strongly agree): (1) I believe that the actions mentioned in the policy plan will assist in fulfilling the policy goal; (2) I believe that the actions mentioned in the policy plan were designed in a professional manner; (3) I believe that the policy plan is in the interest of citizens; (4) I believe that the policy plan reflects a genuine attempt to improve the well-being of citizens; (5) I believe that EPM made an honest attempt to design a good policy plan; (6) I believe that the Ministry of Environmental Protection aims to keep its commitments in that policy plan. The mean of this trust index is `r SVIVA2_01_comb%>%summarise(mean(trust))%>%max()%>%round(2)`, the standard Deviations is `r SVIVA2_01_comb%>%summarise(sd(trust))%>%max()%>%round(2)` and the median is `r SVIVA2_01_comb%>%summarise(median(trust))%>%max()%>%round(2)`. Higher levels of trust are associated with higher trust in government, left-wing ideology and lower levels of education. In addition to measuring participants trust in each policy, I also calculated the delta of each subject’s trust in the strong and weak policy plans that were presented to them (*Mean* = `r SVIVA2_01%>%summarise(mean(WITHIN_DELTA))%>%max()%>%round(2)`, *SD* = `r SVIVA2_01%>%summarise(sd(WITHIN_DELTA))%>%max()%>%round(2)`, *Median* = `r SVIVA2_01%>%summarise(median(WITHIN_DELTA))%>%max()%>%round(2)`). I use this Trust in policy delta to analyze the interaction between symbols and information within subjects.

@@

A growing body of Public-Administration research seeks to assess whether and how government organizations can increase citizens' trust through direct communications with citizens. This study contributes to this discussion by pointing to the influence of symbolic elements in public communications, and unpacking the theoretical mechanisms underlying their effect. Using a rigorous experimental design, embedded in a realistic setting of the Israeli Environmental Protection Ministry, I tested the effect of symbolic elements on trust, and its interactions with the substantive content of communications. To the best of my knowledge, this is the first study to explore this interaction in the context of government communications.

I found that participants' exposure to the familiar symbols of EPM increased their trust in its policy plans, and that the effect was greater when they were presented with a weak policy plan. The latter interaction is partly explained by the negative effect of the symbols on the elaboration of the policy plans. These findings are based on analyses of the responses of more than `r (SVIVA2_01%>%nrow()%>%round(-2))-100` participants to two different policies, and are robust for various model specifications. Moreover, they are evidenced in both between- and within-subjects analyses. These results, if generalizable, suggest that incorporating familiar symbolic elements in public communications increase citizens' trust in government organizations and their actions, via two mechanisms: First, they evoke general positive feelings that are transferred to the specific organizations and policies. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thereby, they cause them to disregard its weaknesses. These two complementary mechanisms are theorized using the framework of the social psychology elaboration likelihood model [@petty_1986; @petty_2011].

Still, one limitation on the experimental findings regards the effect of the symbols on citizens' trust in the strong policy plan, which was relatively weak and marginally significant (Regression model 1.3). One tentative explanation for this could be that symbolic elements are more effective in shifting people from low to medium levels of trust, than shifting them from medium to high levels. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust. If this explanation holds, then result will be a ceiling effect. This explanation can be also linked, theoretically, to the notion of public sector negative bias, which is well established in the Public Administration. Since most citizens tend to have implicit negative attitudes towards the public sector, they can trust public organizations only to a certain degree.[^footnote_ceiling_effect]

@@

##Appendix A: Information manipulation

@@

##Appendix B: Symbols manipulation
![](C:/SAAR/UNIVERSITY/R/SVIVA/papers/paper2_JPART/revision/myfigures/symbols_manipulation.png)

>>>>

---
title: "Can government public communications elicit undue trust? Exploring the interaction between symbols and substantive information in communications"
author: '**Saar Alon-Barkat**, Ph.D.; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  word_document: default
  pdf_document:
    toc: yes
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
link-citations: yes
bibliography: phd_paper_2.bib
csl: ajps.csl
urlcolor: blue
---

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. Recent studies pointed to the emotional effect of symbolic elements, entangled in government public communications (e.g. logos, images and celebrities). Still, they did not examine the interaction between these symbols and the substantive information in communications about bureaucracies’ performance and policies. Exploring this interaction is important for understanding the theoretical mechanisms underlying the effect of symbolic communication on citizens’ trust. Also, it is essential to assess symbols' potency to unduly compensate for unfavorable or logically unpersuasive information, and enable public organizations to escape justified public criticism. Building on the social psychology Elaboration Likelihood Model, I theorize that symbols may increase citizens' trust by conducing citizens to pay less attention to logically unpersuasive information, and thus offsetting its negative effect. I test this indirect mechanism via a large survey experiment, focusing on the Israeli Environment Protection Ministry. The experimental results support the research hypotheses and suggest that the effect of symbolic elements is stronger when communications include logically unpersuasive information. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

One of the main remedies proposed for this concern is greater investment in direct government communications to citizens. Congruently, over the past generation, economically-developed democratic countries have witnessed an increase in government public communications, specifically through advertising, social media, and e-government platforms [@eshuis_2012; @marland_2017; @mergel_2013; @mickoleit_2014; @canel_2013]. In the US, for example, federal government agencies spend approximately 1.5 Billion dollars annually on contracts related to public relations and advertising activities in traditional and digital media [@unitedstatesgovernmentaccountabilityoffice_2017].

Extant research on the effect of government public communications on citizens trust has mostly focused on the *informative* aspects of these communications, whereby government organizations provide citizens with evidence about their performance and details and explanations about their policies and actions [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @porumbescu_2018; @grimmelikhuijsen_2015; @ho_2017]. This focus on information and transparency is in line with the positive normative view of government public communication and public relations, as articulated by the public administration literature on this topic, led by the works of Merdechai Lee. Communications with citizens are perceived as a means for accomplishing democratic accountability, facilitating the effective implementation of their missions, and aligning citizens' perceptions of public organizations with their actual performance [@canel_2013; @lee_2000; @lee_2012; @lee_2001; @lee_2008; @lee_2011]. Importantly, according to this view, public communications are expected to enhance citizens' trust in bureaucracies mainly insofar as these communications provide citizens with favorable information on the performance of organizations and logically persuasive explanations for their policies.

Recently, however, a few studies pointed to the additional effect of *symbolic* elements that are entangled in government communications, such as  agencies' names and brand logos, figures, images and celebrity endorsements [@alonbarkat_2017; @karens_2016; @teodoro_2018]. These studies demonstrated that employing familiar and well-designed symbolic elements that evoke positive associations and emotions may cause citizens to view government organizations more favorably and trust them. They suggested that the influence of symbols occurs unconsciously through a psychological mechanism of transfer of affect or "evaluative conditioning".

The distinction between the effects of symbolic elements and substantive information in public communications has significant normative implications. Whereas the responsiveness of citizens to substantive information is perceived favorably from a democratic standpoint (assuming that the information is accurate indeed), establishing trust through symbolic elements can be highly problematic. Shaping citizens' perceptions of governments by evoking associations and emotions stands against the normative expectation that citizens' judgments should be based on their critical thinking. Furthermore, we need to be aware that symbolic communication may elicit undue trust in poorly performing organizations and in ill-conceived policies, and enable public organizations to escape justified public criticism [@alonbarkat_2017].

Whilst previous studies explored the distinct effect of symbolic elements on trust, they did not examine its interaction with the substantive information in the communication. Exploring this latter interaction is important for understanding the theoretical mechanisms underlying the effect of symbolic communication on citizens’ trust. Also, it is essential to assess the possibility that symbols can attenuate the effect of substantive information on trust, and specifically - to attenuate the rational negative responses of citizens to unfavorable or logically unpersuasive information.

To theoretically address the latter research gap, I draw upon the social psychology Elaboration Likelihood Model (ELM), which provides a comprehensive theoretical framework for various psychological processes underlying individuals' responses to persuasive communication, and their consequences for attitude change [@petty_1986; @petty_2011]. Based on ELM, I suggest that symbolic elements in public communications can affect citizens' trust not only through evaluative conditioning, as suggested by previous studies, but also by conducing citizens to pay less attention to logically unpersuasive information that could otherwise decreased their trust. According to the latter mechanism, symbols are expected to be more effective when used in communications that include logically unpersuasive information.

To empirically test these theoretical expectations, I conducted a large randomized survey experiment among circa `r SVIVA2_01%>%nrow()%>%round(-1)` Israeli citizens. The experiment focuses on the empirical case of the Israeli Environment Protection Ministry. I explore citizens' responses to public communications about two policy plans of the Ministry. Specifically, I examine how their trust in these policies is affected by the symbolic elements in the communications (real symbols vs. fake symbols and no symbols conditions) and its interaction with the substantive information in them (logically persuasive vs. unpersuasive policy plans).

I found that the symbolic elements increased citizens' trust in these policies. In line with my theoretical expectations, the effect was greater among those who saw the logically unpersuasive policy plans. The latter were less inclined to think about the weak, unpersuasive content, since they were distracted by the symbols. I discuss the theoretical and normative implications of these findings.

@@

A symbol can be broadly defined as *"any object used by human beings to index meanings that are not inherent in, nor discernible from, the object itself"* [@elder_1983, pp. 28-29]. Symbols are fundamental for almost any form of human interaction, and specifically in the context of politics and public affairs [@edelman_1964; @edelman_1988; @elder_1983]. Constellations of symbols in discourse form consolidated narratives, which play an key role in shaping people's perceptions of policy problems and solutions [@miller_2012; @miller_2015].

My focus in this study is on those symbols that are strategically used by government organizations in their public communications, and designed to distinguish them and to evoke associations and emotions that stimulate audiences’ positive affect towards organizations and their operations. Prominent examples are unique logos, colors associated with specific meanings, icons, images, and the use of celebrities as endorsers. These elements are widely used by businesses and political candidates to attract targeted consumers and voters. Yet, they can also be found in the communications of government bureaucracies with citizens, including their mass-media advertising campaigns, social media accounts, internet pages, official publications, signboards in public places and personal mail notifications to citizens.

@@

@alonbarkat_2017 examined Israeli citizens' trust in the Israel Electricity Corporation, a state owned company, and it's shaping by their exposure to comic cartoon figures that were routinely presented in the organization's public campaigns. They compared citizens who were personally exposed to prolonged power outages and others who were not. They found that the symbolic element had a positive effect on citizens' trust in the organization, including among those who personally experienced poor performance (a modest effect size equivalent to 0.25 standard deviations). Building on social psychology theory and marketing research, they suggested that the symbols enhance trust by inducing an unconscious transfer of affect from the symbols to the organizations, after repeated linkage between them ("evaluative conditioning"). They proposed that the effectiveness of government symbolic elements depends on their familiarity, and aesthetic design, as well as on the ambiguity of the government quality assessed.

@@

Even though these studies substantiate a link between the appearance of symbolic elements in communications and citizens' greater trust in public organizations and policies, they neither theorized nor empirically tested the interaction between symbolic elements and citizens’ processing of substantive information in communications. We do not know whether well-designed symbols reinforce, or offset, citizens' responses to information. To address this theoretical gap in public administration research, I propose a nuanced theoretical framework for the effects of symbols in communications and their interaction with information, building on social psychology theory on persuasive communication.

@@

The social psychology Elaboration Likelihood Model (ELM) provides a comprehensive theoretical framework for investigating peoples' varying responses to "persuasive" communications [@petty_1986; @petty_2011]. It theorizes when and how individuals are more/less likely to change their attitudes following a communication or a message. Accordingly, ELM can be applied to various settings, types of attitudes and forms of communications. The theory has been applied mainly in relation to commercial advertising and marketing [for a review see: @schumann_2012], but also to other communication settings, including public health campaigns [@petty_2009; @briol_2006], media coverage [@petty_2002; @vanderwurff_2016] and entertainment [@young_2008; @slater_2002].[^footnote_elm_critique] So far however the relevance of ELM has not been directly examined in relation to changes in citizens' trust in public organizations in response to their public communications.

[^footnote_elm_critique]: For a summary of the ELM literature, and its main critique, see: @kitchen_2014.

According to ELM, the responses of individuals to persuasive communications depends on how much effort they devote to thinking about the content of the message (or "the amount of thinking" according to ELM terminology). Persuasion, or attitudinal change, can occur when thinking is "high" or "low", but the processes and consequences are likely to differ in each situation. This range between high and low degrees of thought is referred to as the "elaboration continuum". People's location along this continuum is associated with different psychological processes of persuasion. At the high end of the elaboration continuum, also called the "central route", people are likely to respond by cognitively scrutinizing the arguments central to the merits of the issue. At the low end, also called "the peripheral route" people respond to simple cues automatically and unconsciously through primitive psychological processes that require very little cognitive effort. One of these latter processes is evaluative conditioning.[^footnote_dual_process_theories] In turn, these two mechanisms are likely to cause divergent types of persuasion. Persuasion via the central route is expected to be more enduring, resistant and predictive of behavior, compared with the peripheral route. Of course, often people are located in between the two ends of the continuum, and thus persuasion is determined by a combination of these processes.

@@

In light of the conceptualization of ELM, substantive information elements in government public communications represent issue-related arguments, which can be scrutinized or processed systematically based on their logical quality (via the central route), at least by those who are sufficiently motivated and able to do so. More positive information about organizations' performance and more logically persuasive arguments supporting their actions would increase citizens' trust in these organizations and their actions and vice versa. Conversely, symbolic elements are potential peripheral cues. They can increase citizens' trust via the peripheral route, through evaluative conditioning. Importantly, in addition to their direct effect as peripheral cues, symbolic elements may also increase citizens' trust indirectly. They can decrease the amount of thought citizens dedicate to the communication, and thus prevent them from scrutinizing substantive information that would otherwise decrease their trust, namely unfavorable information (e.g. information on poor performance) and weak arguments (e.g. logically unpersuasive arguments supporting their actions and policies). The latter assertion entails an interaction between symbols and the substantive information in communications: symbolic elements in public communications are more likely to affect citizens' trust when communications include unfavorable information or weak logical arguments.[^footnote_interaction_strong_information]

@@

To empirically test the above mentioned hypotheses, I designed a randomized survey experiment, which simulates the effects of symbols and information in government organizations' communications about their policy plans. I focus on communication of future policy plans, as opposed to communications on performance of existing policies. This allows me to utilize a real-world empirical case, yet still to manipulate the content of the communication in a reliable manner. Most citizens do not have prior knowledge of organizations' future policy plans, and will therefore be more willing to trust a fictional communication of an organization with regard to its future policy.

@@

The empirical setting of this experiment relates to the Israeli Environment Protection Ministry (hereafter: EPM) that is responsible for the formulation, co-ordination and execution of environmental policy in Israel, at the national and local levels. I analyzed the effects of informative and symbolic elements in the EPM's communications of two policy plans on citizens' trust in these policies, and the extent to which they elaborated on them. In the survey, the communications about the two policy plans are presented to the participants as sections from the EPM's official publication of its annual work plan.

@@

In the course of the survey experiment, I was able to confirm that the EPA logo and the images of the celebrities from the campaigns were familiar to the participants. A total of `r round((nrow(filter(SVIVA2_01,RECOGNIZE_SVIVA_logo>1))/N_1)*100,0)`% of the participants reported that they recognize the EPA logo to some extent, and a vast majority of them reported that they recognize the campaign images of Friedman and Rosenblum (`r round((nrow(filter(SVIVA2_01,RECOGNIZE_air_real>1))/N_1)*100,0)`% and `r round((nrow(filter(SVIVA2_01,RECOGNIZE_waste_real>1))/N_1)*100,0)`%, respectively). In robust analyses I further exclude those who do not recognize these symbols. Additionally, in two prior surveys, I collected further information about the affect and symbolic associations of these symbols. The results of these tests, reported in the supplementary appendix, substantiate my empirical assumption that these elements evoke positive emotions, and that citizens automatically associate them with environmental protection policy and the Ministry.

@@

@@

In the control no-symbols condition, the two policies were displayed in a minimal black and white design, and without the logo and the celebrities’ images. In the control fake-symbols condition, the green color was replaced with blue, the EPM logo was replaced by a fake logo and the images of celebrities from the campaigns were replaced with two edited images of unfamiliar people. The purpose of the fake-logo condition was to isolate the symbolic meaning of the EPM's graphic identifiers and distinguish it from the possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating emotions and associations which could be attached to the government ministry and environmental issues. In that sense, the "fake" symbols could be regarded as a placebo. In two prior surveys, I confirmed, empirically, that the fake symbols are unfamiliar, do not have strong positive or negative affect, and are not strongly associated with environment policy and EPM. The results of these examinations are reported in the supplementary appendix.

@@

@@

The main independent variables are the above mentioned manipulations for information and symbols, and the interaction between them. The outcome variable is citizens' trust in policy (or the perceived trustworthiness of the policy). Trust in policy is also closely related to the concept of policy credibility [@gilardi_2002; @greasley_2014]. It is measured in the survey for each policy plan using a composite index of six Likert scale items (Cronbach’s alpha = `r alpha_waste_trust`), which were adopted and slightly modified from validated trust scales used by previous public administration studies on trust in government [@grimmelikhuijsen_2013; @grimmelikhuijsen_2014; @karens_2016; @porumbescu_2016; @grimmelikhuijsen_2017]. Participants were asked to indicate their agreement with the following statements, between 1 (weakly agree) and 7 (strongly agree): (1) I believe that the actions mentioned in the policy plan will assist in fulfilling the policy goal; (2) I believe that the actions mentioned in the policy plan were designed in a professional manner; (3) I believe that the policy plan is in the interest of citizens; (4) I believe that the policy plan reflects a genuine attempt to improve the well-being of citizens; (5) I believe that EPM made an honest attempt to design a good policy plan; (6) I believe that the Ministry of Environmental Protection aims to keep its commitments in that policy plan. The mean of this trust index is `r SVIVA2_01_comb%>%summarise(mean(trust))%>%max()%>%round(2)`, the standard Deviations is `r SVIVA2_01_comb%>%summarise(sd(trust))%>%max()%>%round(2)` and the median is `r SVIVA2_01_comb%>%summarise(median(trust))%>%max()%>%round(2)`. Higher levels of trust are associated with higher trust in government, left-wing ideology and lower levels of education. In addition to measuring participants trust in each policy, I also calculated the delta of each subject’s trust in the strong and weak policy plans that were presented to them (*Mean* = `r SVIVA2_01%>%summarise(mean(WITHIN_DELTA))%>%max()%>%round(2)`, *SD* = `r SVIVA2_01%>%summarise(sd(WITHIN_DELTA))%>%max()%>%round(2)`, *Median* = `r SVIVA2_01%>%summarise(median(WITHIN_DELTA))%>%max()%>%round(2)`). I use this Trust in policy delta to analyze the interaction between symbols and information within subjects.

@@

A growing body of Public-Administration research seeks to assess whether and how government organizations can mitigate citizens' distrust in them through direct communications with citizens. This study contributes to this discussion by theorizing and testing the direct and indirect mechanisms through which symbolic elements in communications can affect citizens' trust. Using a rigorous experimental design, embedded in a realistic setting of the Israeli Environmental Protection Ministry, I tested the interaction between symbolic elements and the substantive content of communications. To the best of my knowledge, this is the first study to explore this interaction in the context of government communications.

I found that participants' exposure to the familiar symbols of EPM increased their trust in its policy plans, compared with two control groups, and that the effect was greater when participants were presented with a weak, logically unpersuasive, policy plan. This interaction between symbols and information is explained, in part, by the negative effect of the symbols on participants' elaboration of the information on policy plans. In other words, the symbols distracted participants from scrutinizing the content of the communication, thus offsetting its negative impact. These findings are based on analyses of the responses of more than `r (SVIVA2_01%>%nrow()%>%round(-2))-100` participants to two different policies, and are robust for various model specifications. Moreover, they are evidenced in both between- and within-subjects analyses. These results, if generalizable, suggest that incorporating familiar symbolic elements in public communications increase citizens' trust in government organizations and their actions, via two mechanisms: First, symbols evoke general positive feelings that are transferred to the specific organizations and policies. Second, by putting citizens in a positive mood, they distract them from thinking about the substantive content of the communication, and thereby, they cause them to disregard its weaknesses. These two complementary mechanisms are theorized using the framework of the social psychology elaboration likelihood model [@petty_1986; @petty_2011].

Still, one limitation on the experimental findings regards the effect of the symbols on citizens' trust in the strong policy plan, which was relatively weak and marginally significant (Regression model 1.3). One tentative explanation for this could be that symbolic elements are more effective in shifting people from low to medium levels of trust, than shifting them from medium to high levels. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust. This explanation can be also linked, theoretically, to the notion of public sector negative bias [@hvidman_2018; @hvidman_2016; @marvel_2015a]. Since many citizens tend to have implicit negative attitudes towards the public sector, they can trust public organizations only to a certain degree.[^footnote_ceiling_effect] Still, the latter tentative explanation merits further empirical testing in future research.

@@

#Appendix A: Information manipulation

@@

#Appendix B: Symbols manipulation
![](C:/SAAR/UNIVERSITY/R/SVIVA/papers/paper2_JPART/revision/myfigures/symbols_manipulation.png)
